---
title: "Moran Eigenvectors"
author: "Roger Bivand"
output:
  html_document
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{Moran Eigenvectors}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Introduction

The Moran eigenvector approach
[@dray+legendre+peres-neto:06; @griffith+peres-neto:06] involved the
spatial patterns represented by maps of eigenvectors; by choosing
suitable orthogonal patterns and adding them to a linear or generalised
linear model, the spatial dependence present in the residuals can be
moved into the model.

It uses brute force to search the set of eigenvectors of the matrix
$\mathbf{M W M}$, where

$$\mathbf{M} = \mathbf{I} - \mathbf{X}(\mathbf{X}^{\rm T}
\mathbf{X})^{-1}\mathbf{X}^{\rm T}$$ is a symmetric and idempotent
projection matrix and $\mathbf{W}$ are the spatial weights. In the
spatial lag form of [`SpatialFiltering`]{} and in the GLM [`ME`]{} form
below, $\mathbf{X}$ is an $n$-vector of ones, that is the intercept
only.

In its general form, [`SpatialFiltering`]{} chooses the subset of the
$n$ eigenvectors that reduce the residual spatial autocorrelation in the
error of the model with covariates. The lag form adds the covariates in
assessment of which eigenvectors to choose, but does not use them in
constructing the eigenvectors. [`SpatialFiltering`]{} was implemented
and contributed by Yongwan Chun and Michael Tiefelsdorf, and is
presented in @tiefelsdorf+griffith:07; [`ME`]{} is based on Matlab code
by Pedro Peres-Neto and is discussed in @dray+legendre+peres-neto:06 and
@griffith+peres-neto:06.

```{r echo=TRUE}
library(spdep)
if (require(rgdal, quietly=TRUE)) {
  NY8 <- readOGR(system.file("shapes/NY8_utm18.shp", package="spData"))
} else {
  require(maptools, quietly=TRUE)
  NY8 <- readShapeSpatial(system.file("shapes/NY8_utm18.shp", package="spData"))
}
NY_nb <- read.gal(system.file("weights/NY_nb.gal", package="spData"), region.id=row.names(NY8))
```

```{r echo=TRUE,eval=TRUE}
nySFE <- SpatialFiltering(Z~PEXPOSURE+PCTAGE65P+PCTOWNHOME, data=NY8, nb=NY_nb, style="W", verbose=FALSE)
nylmSFE <- lm(Z~PEXPOSURE+PCTAGE65P+PCTOWNHOME+fitted(nySFE), data=NY8)
summary(nylmSFE)
nylm <- lm(Z~PEXPOSURE+PCTAGE65P+PCTOWNHOME, data=NY8)
anova(nylm, nylmSFE)
```

Since the [`SpatialFiltering`]{} approach does not allow weights to be
used, we see that the residual autocorrelation of the original linear
model is absorbed, or ‘whitened’ by the inclusion of selected
eigenvectors in the model, but that the covariate coefficients change
little. The addition of these eigenvectors – each representing an
independent spatial pattern – relieves the residual autocorrelation, but
otherwise makes few changes in the substantive coefficient values.

The [`ME`]{} function also searches for eigenvectors from the spatial
lag variant of the underlying model, but in a GLM framework. The
criterion is a permutation bootstrap test on Moran’s $I$ for regression
residuals, and in this case, because of the very limited remaining
spatial autocorrelation, is set at $\alpha = 0.5$. Even with this very
generous stopping rule, only few eigenvectors are chosen; their combined
contribution only just improves the fit of the GLM model.

```{r echo=TRUE}
NYlistwW <- nb2listw(NY_nb, style = "W")
set.seed(111)
```

```{r echo=TRUE,eval=FALSE}
nyME <- ME(Cases~PEXPOSURE+PCTAGE65P+PCTOWNHOME, data=NY8, offset=log(POP8), family="poisson", listw=NYlistwW, alpha=0.5)
```

```{r echo=FALSE, eval=FALSE}
bsfn <- system.file("etc/backstore/nyME_res.RData", package="spdep")
load(bsfn)
```

```{r echo=TRUE, eval=FALSE}
nyME
NY8$eigen_24 <- fitted(nyME)[,1]
NY8$eigen_223 <- fitted(nyME)[,2]
```

```{r results='asis',echo=FALSE, eval=FALSE}
.iwidth <- 6
.iheight <- 4
.ipointsize <- 10
library(RColorBrewer)
#gry <- brewer.pal(9, "Greys")[-1]
spplot(NY8, c("eigen_24", "eigen_223"), col.regions=grey.colors(6, 0.95, 0.55, 2.2), cuts=5)
``` 

```{r echo=TRUE,eval=FALSE}
nyglmME <- glm(Cases~PEXPOSURE+PCTAGE65P+PCTOWNHOME+offset(log(POP8))+fitted(nyME), data=NY8, family="poisson")
summary(nyglmME)
nyGLMp <- glm(Cases~PEXPOSURE+PCTAGE65P+PCTOWNHOME+offset(log(POP8)), data=NY8,family="poisson")
anova(nyGLMp, nyglmME, test="Chisq")
```

Figure \[fig:geigen2\] shows the spatial patterns chosen to match the
very small amount of spatial autocorrelation remaining in the model. As
with the other Poisson regressions, the closeness to TCE sites is highly
significant. Since, however, many TCE sites are also in or close to more
densely populated urban areas with the possible presence of both
point-source and non-point-source pollution, it would be premature to
take such results simply at their face value. There is, however, a
potentially useful contrast between the cities of Binghampton in the
south of the study area with several sites in its vicinity, and Syracuse
in the north without TCE sites in this data set.

## References

[^1]: This vignette formed pp. 302–305 of the first edition of Bivand,
    R. S., Pebesma, E. and Gómez-Rubio V. (2008) Applied Spatial Data
    Analysis with R, Springer-Verlag, New York. It was retired from the
    second edition (2013) to accommodate material on other topics, and
    is made available in this form with the understanding of the
    publishers.
